Reagents: expressing and composing fine-grained concurrency

### 2011-10-13 paper/paper.tex
added some macros for semantics
progress on paper

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{mathpartir}

\newcommand{\ra}{\rightarrow}
\newcommand{\gives}{\vdash}
\newcommand{\proves}{\vdash}
\newcommand{\GA}{\ |\ }
\newcommand{\kw}[1]{\textsf{#1}}
\newcommand{\ty}[1]{\textrm{#1}\ }
\newcommand{\bind}{\mbox{\ \scriptsize$>\!\!>\!=$\ }}
\newcommand{\letrec}[2]{\kw{letrec}\ #1\ \kw{in}\ #2}
\newcommand{\LET}[4]{\kw{let}_{#1}\ #2 = #3\ \kw{in}\ #4}

@@

\authorinfo{}{}{}
% yikes!
\title{Reagents: expressing and composing patterns of fine-grained concurrency}
\maketitle

\begin{abstract}
blah blah
\end{abstract}

@@

Such libraries are an enormous undertaking---and one that must be repeated for
new platforms.  They tend to be conservative, implementing only those data
structures and primitives likely to fulfill common needs.  In addition, it is
generally not possible to safely combine the facilities of the library.  For
example, JUC provides queues, sets and maps, but not stacks or bags.  Its
queues come in both blocking and nonblocking forms, while its sets and maps
are nonblocking only.  Although the queues provide atomic (thread-safe)
dequeuing and sets provide atomic insertion, it is not possible to combine
these into a single atomic operation.

In short, libraries for fine-grained concurrency are indispensable, but hard
to write, hard to extend by composition, and hard to tailor to the exact needs
of a particular user.

@@

We have developed a core library of \emph{composable} primitives for
fine-grained concurrency.

@@

why is this different from STM

@@

@@

@@

@@

\bibliographystyle{abbrvnat}

@@

# Contributions

  - generalizes transactions events, but with much better
  implementation

@@

2.5 Title, abstract, intro, library overview 
1.5 Semantics 
2.5 Implementation/algorithms 
1.0 Performance results
1.5 Related work
1.0 Conclusion and bib

>>>>

\newcommand{\elide}[1]{}

\input{packages}

@@

\authorinfo{}{}{}
% yikes!
%\title{Reagents: expressing and composing patterns of fine-grained concurrency}
\title{Reagents: expressing and composing fine-grained concurrency}
\maketitle

\begin{abstract}
blah blah\\
blah blah\\
blah blah\\
blah blah\\
blah blah\\
blah blah
\end{abstract}

@@

Such libraries are an enormous undertaking---and one that must be repeated for
new platforms.  They tend to be conservative, implementing only those data
structures and primitives likely to fulfill common needs.  In addition, it is
generally not possible to safely combine the facilities of the library.  For
example, JUC provides queues, sets and maps, but not stacks or bags.  Its
queues come in both blocking and nonblocking forms, while its sets and maps
are nonblocking only.  Although the queues provide atomic (thread-safe)
dequeuing and sets provide atomic insertion, it is not possible to combine
these into a single atomic operation that moves an element from a queue into a
set.

In short, libraries for fine-grained concurrency are indispensable, but hard
to write, hard to extend by composition, and hard to tailor to the needs of
particular users.

@@

% first class synchronous operations

%% We have developed a small set of primitives for fine-grained concurrency that
%% are \emph{expressive} and \emph{composable}:

%% We have developed \emph{reagents}, a new abstraction for fine-grained
%% concurrency that is \emph{expressive} and \emph{composable}:

%% We have developed \emph{reagents}, which are first-class representations of
%% fine-grained concurrency primitives.  Reagents are \emph{expressive} and
%% \emph{composable}:

We have developed \emph{reagents}, which abstractly represent fine-grained
concurrent operations.  Reagents are \emph{expressive} and \emph{composable},
but retain the scalability and performance of existing algorithms:

%We have developed \emph{reagents},

%% compositional primitives for fine-grained concurrency

%% a small set of primitives for fine-grained concurrency that
%% are \emph{expressive} and \emph{composable}:

\paragraph{Expressive}  .

\paragraph{Composable} .

We see this work as serving the needs of two distinct groups: concurrency
experts and concurrency users.  Using reagents, experts can write algorithms
more easily, because common patterns are expressible as abstractions and many
are built-in.  Users can compose, extend, and tailor the library without
detailed knowledge of the algorithms involved.

blend shared-state and message-passing concurrency

\elide{
# Contributions

  - generalizes transactional events, but with more specialized
  implementation (doesn't need an underlying STM)

@@

why is this different from STM
}

\section{Background}

Reagents are directly motivated by

Before introducing reagents, it will be helpful to

readers familiar with these algorithms can safely jump to sec:reagents

also background on Scala

traditional code for Treiber stack (with exp. backoff),

\begin{lstlisting}
class TreiberStack[A] {
  private val head = new AtomicReference[List[A]](Nil)
  def push(a: A) {
    val backoff = new Backoff
    while (true) {
      val cur = head.get()
      if (head.cas(cur, a :: cur)) return
      backoff.once()
    }
  }
  def tryPop(): Option[A] = {
    val backoff = new Backoff
    while (true) {
      val cur = head.get() 
      cur match {
        case Nil     => return None
        case a::tail => if (head.cas(cur, tail)) return a
      }
      backoff.once()
    }
  }
}
\end{lstlisting}

need to explain pattern matching and PartialFunction

talk about elimination backoff and blocking versions but don't show code.

discuss the insight behind MCS locks

@@

First we give an overview of the library, and then we illustrate its
expressiveness and composability by showing how to implement several

chemistry metaphor

lack of isolation needed to express many lock-free algorithms.
message-passing turns out to be a good way to understand what's going on.

\begin{lstlisting}
class TreiberStack[A] {
  private val head = new Ref[List[A]](Nil)
  val push: Reagent[A, Unit] = head.upd {
    (xs, x) => (x:xs, ())
  }
  val tryPop: Reagent[Unit, Option[A]] = head.upd {
    case (x::xs, ()) => (xs, Some(x))
    case (Nil,   ()) => (Nil, None)
  }
}
\end{lstlisting}

Examples:

\begin{itemize}
\item TreiberStack
\item with blocking
\item with elimination backoff
\item MSQueue
\item with blocking
\item with buckets?
\item Counter -> semaphore
\item RWLocks
\item Tree barriers
\item Dining philosophers(ish)
\item Flat combining
\end{itemize}

@@

\newcommand{\ra}{\rightarrow}
\newcommand{\gives}{\vdash}
\newcommand{\proves}{\vdash}
\newcommand{\kw}[1]{\textsf{#1}}
\newcommand{\GA}{\ |\ }
\newcommand{\step}{\longrightarrow}
\newcommand{\cstep}[1]{\stackrel{#1}{\Longrightarrow}}

\newcommand{\evalCtx}{\mathbb{E}}
\newcommand{\procCtx}{\mathbb{P}}

\newcommand{\ret}[1]{\kw{ret}(#1)}
\newcommand{\never}{\kw{never}}
\newcommand{\postCommit}[1]{\kw{postCommit}(#1)}
\newcommand{\computed}[1]{\kw{computed}(#1)}
\newcommand{\compose}{{\small\texttt{>=>}}}
\newcommand{\lift}[1]{\kw{lift}(#1)}
\newcommand{\send}[1]{#1}
\newcommand{\cas}[3]{\kw{cas}(#1,#2,#3)}
\newcommand{\rd}[1]{\kw{read}(#1)}
\newcommand{\newChan}{\kw{new Chan}}
\newcommand{\newRef}[1]{\kw{new Ref(#1)}}

\elide{
\[
\infer
  {P_i, \sigma_i \cstep{t_i} P'_i, \sigma'_i}
  {P_1 | \cdots | P_n,\ \sigma_1 \uplus \cdots \uplus \sigma_n\ \step\
   P'_1 | \cdots | P'_n,\ \sigma'_1 \uplus \cdots \uplus \sigma'_n
  }
\]

\[
\infer
  {r_i!v_i, \sigma_i \cstep{t_i} \ret{v'_i}, \sigma'_i}
  {\procCtx[\prod \evalCtx_i[r_i ! v_i]],\ \biguplus \sigma_i \step
   \procCtx[\prod \evalCtx_i[\ret{v'_i}]],\ \biguplus \sigma'_i}
\]
}

\[
\infer
  {r_i!v_i, \sigma_i \cstep{t_i} \ret{v'_i}, \sigma'_i}
  {\prod \evalCtx_i[r_i ! v_i],\ \biguplus \sigma_i \step
   \prod \evalCtx_i[\ret{v'_i}],\ \biguplus \sigma'_i}
\]

@@

\begin{lstlisting}
def tryReact(a: A, rx: Reaction, offer: Offer[B]): Any = 
  r1.tryReact(a, rx, offer) match {
    case Retry => 
      r2.tryReact(a, rx, offer) match {
        case Retry => Retry
        case Block => Retry // must retry r1
        case ans   => ans
      }
    case Block => r2.tryReact(a, rx, offer)
    case ans => ans
  }
\end{lstlisting}

@@

\bibliographystyle{abbrvnat}
\bibliography{reagents}

@@

@@

1.2 Title, abstract, intro, 
1.0 Background
2.5 Library overview & examples
1.5 Semantics 
1.5 Implementation/algorithms 
0.5 Performance results
1.0 Related work
0.8 Bib
